Cubical Types with Positive and Negative Polarities

### 2014-05-19 ring.tex
separate syntax from denotation

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\tau$ for cubical
types: these have the same formal structure as regular types but also include
\emph{negative} types. The syntax of the types is summarized below:
\[\begin{array}{rcl}
t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\tau &::=& t \alt 
      \alt \tau_1 + \tau_2
      \alt \tau_1 * \tau_2
      \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.

It is easy to find a model for regular types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \bullet \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets.

For cubical types, their denotations are $n$-dimensional cubical sets with
polarities, written as $\nodet{\cubt_1}{\cubt_2}$ An $n$-dimensional cubical
set $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 1-dimensional case reduces to plain
sets (which are the denotations of first-order types). As we explain below,
cubical sets are closed under sums, differences, products.

Examples of 1-dimensional cubical sets include all the finite sets $S$
denoting regular first-order types $t$. A 2-dimensional cubical set looks
like $\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1
- t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
3-dimensional cubical type looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
sign as follows: white in the outermost box corresponds to a positive
polarity and gray in the outermost box corresponds to a negative polarity;
then, as we move inwards the polarities alternate for every change of
color.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \den{t} = \nodet{\den{t}}{0} = 
  \nodet{\nodet{\den{t}}{0}}{0} = \ldots \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical types not be balanced in general but it is always
possible to extend the shorter subtree to an arbitrary depth using the first
clause which formalizes that $t$ is equivalent to $t-0$ and to $(t-0)-0$ etc.
In the recursive definitions below, we assume that the types have been
extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~\tau &=& \tau \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

>>>>

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\tau$ for cubical
types: these have the same formal structure as regular types but also include
\emph{negative} types. The syntax of the types is summarized below:
\[\begin{array}{rcl}
t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\tau &::=& t \alt 
      \alt \tau_1 + \tau_2
      \alt \tau_1 * \tau_2
      \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$. At
the level of cubical types, we can write types such as
$(t_1-t_2)-(t_3-t_4*t_5)$ which mix negative types with the usual type
constructors.

It is easy to find a model for regular types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \bullet \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets. We usually use $S$ for the denotation of a
conventional first-order type.

For cubical types, their denotations, usually written as $\cubt$, are
$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 1-dimensional case reduces to
$\nodet{S}{\emptyset}$ which is essentially the denotation of conventional
first-order types. As we explain below, cubical sets are closed under sums,
differences, products.

Examples of 1-dimensional cubical sets include all the sets of the form
$\nodet{S}{\emptyset}$ where $S$ is finite set denoting a regular first-order
types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
sign as follows: white in the outermost box corresponds to a positive
polarity and gray in the outermost box corresponds to a negative polarity;
then, as we move inwards the polarities alternate for every change of color.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \nodet{\den{t}}{\emptyset} = 
  \nodet{\nodet{\den{t}}{\emptyset}}{\emptyset} = \ldots \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical types not be balanced in general but it is always
possible to embed an $n$-dimensional set into a larger dimension using the
equivalence in the first clause which formalizes that $t$ is equivalent to
$t-0$ and to $(t-0)-0$ etc.  In the recursive definitions below, we assume
that the types have been extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~S &=& S \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]
